---
title: "Homework Problem Set 9: Choosing a Sample Size"
output: 
  html_document: 
    theme: readable
  pdf_document: default
header-includes:
  - \usepackage{array}
  - \usepackage{booktabs}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, out.width = "100%", fig.align = "center", 
  cache = FALSE, dev = ifelse(knitr::is_html_output(), "png", "tikz"))
```

```{r packages}
library(tidyverse)
library(kableExtra)
```

```{r utilities}
source("../../utilities.R")
```

`r ifelse(knitr::is_html_output(), paste("You can also download a [PDF](homework-09.pdf) copy of this homework assignment.", sep = ""), "")`

Note: In each problem I give the answer. You should be able to get the same results.

## Replicating Mendel's Pea Study

Suppose we wanted to replicate Mendel's pea study described in an [earlier set](homework-07.html) of homework problems. The goal is to estimate the probability ($p$) of an offspring with yellow peas using the proportion ($\hat{p}$) of offspring with yellow peas out of a sample of $n$ observations. 

1. Assume we want a confidence level of 95% and a margin of error of 0.05. Confirm that if we use $p$ = 0.75 as a *good guess* of $p$ then the sample size we should use is approximately `r round(1.96^2*0.75*0.25/0.05^2)`, and that the *upper bound* on the sample size we should use is `r round(1.96^2*0.25/0.05^2)`.

0. Assume we want a confidence level of 99% and a margin of error of 0.05. Confirm that if we use $p$ = 0.75 as a *good guess* of $p$ then the sample size we should use is approximately `r round(2.576^2*0.75*0.25/0.05^2)`, and that the *upper bound* on the sample size we should use is `r round(2.576^2*0.25/0.05^2)`.

```{r}
n <- 25
s <- 15
m <- 250
```

## Another Trebuchet Study

Recall the study of the trebuchet at Château des Baux described in a [previous set](homework-07.html) of homework problems. In that study they fired the trebuchet `r n` times and observed a mean distance of `r m` meters with a standard deviation of `r s` meters. This resulted in a margin of error of approximately `r round(qt(0.975, df = n - 1), 3) * s / sqrt(n)` meters. Now suppose we wanted to conduct another study that would produce a smaller margin of error. 

1. Confirm that the sample size necessary for a margin of error of approximately 3 meters with the same confidence level (95%) is approximately `r round(1.96^2 * s^2 / 9)` if we use the sample standard deviation from this previous study as an estimate of $\sigma$. That is, we would need to fire the trebuchet approximately `r round(1.96^2 * s^2 / 9)` times to produce a margin of error of 3 meters.

0. Confirm that the sample size necessary for a margin of error of approximately one meter with the same confidence level (95%) is approximately `r round(1.96^2 * s^2 / 1)` if we use the sample standard deviation from this previous study as an estimate of $\sigma$. That is, we would need to fire the trebuchet approximately `r round(1.96^2 * s^2 / 1)` times to produce a margin of error of one meter.

```{r}
rm(list = ls())
```

