---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("10-01-2022","12-01-2022","14-01-2022") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 1
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction and the "big picture"](lecture-01-12-2022.html).

`r i <- i + 1; d[i]` [Distributions, graphical depictions of distributions (i.e., histograms and dot plots), and summary measures](lecture-01-14-2022.html). A set of [homework problems](homework-01.html) on describing distributions in now available.

***

`r i <- i + 1; d[i]` **Holiday**.

`r i <- i + 1; d[i]` [Summary measures --- continued, cumulative distributions, and box plots](lecture-01-19-2022.html). A set of [homework problems](homework-02.html) on these topics is now available.

`r i <- i + 1; d[i]` [Normal/bell-shaped distributions](lecture-01-21-2022.html).
 
***

`r i <- i + 1; d[i]` [Outlier identification, margin of error](lecture-01-24-2022.html). A set of [homework problems](homework-03.html) on normal distributions and outliers is now available. The [study guide](e1-studyguide.html) for the first examination is also now available.

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **First examination**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]`
 
***

`r i <- i + 1; d[i]` **Second examination**.

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Holiday**.

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **No class**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Third examination**.

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Fourth examination**.

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Fifth examination**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Sixth examination**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

***

`r i <- i + 1; d[i]` **Final examination, 10:15 AM to 12:15 PM**.


