---
output:
  html_document:
    theme: readable
---

```{r, echo = FALSE, message = FALSE}
library(lubridate)
date <- "10-06-2021"
weekday <- wday(mdy(date), label = TRUE, abbr = FALSE)
month <- month(mdy(date), label = TRUE)
day <- day(mdy(date))
```

---
title: `r paste(weekday, ", ", month, " ", day, sep = "")`
header-includes:
  - \usepackage{array}
  - \usepackage{booktabs}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.align = "center", out.width = "100%", fig.width = 9, cache = FALSE, dev = ifelse(knitr::is_html_output(), "png", "pdf"))
```

```{r packages}
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
```

```{r utilities}
source("../../utilities.R")
```

`r ifelse(knitr::is_html_output(), paste("You can also download a [PDF](lecture-", date, ".pdf) copy of this lecture.", sep = ""), "")`

## Estimation Equations

```{r}
parameter <- c("$\\mu$", "$p$", "$\\mu$", "$\\tau$")
estimator <- c("$\\bar{x}$", "$\\hat{p}$", "$\\bar{x}$", "$N\\bar{x}$")
se <- c("$\\frac{s}{\\sqrt{n}}$", "$\\sqrt{\\frac{\\hat{p}(1-\\hat{p})}{n}}$", 
  "$\\frac{s}{\\sqrt{n}}\\sqrt{1-\\frac{n}{N}}$",    "$N\\frac{s}{\\sqrt{n}}\\sqrt{1-\\frac{n}{N}}$")
me <- c("$t\\frac{s}{\\sqrt{n}}$", "$z\\sqrt{\\frac{\\hat{p}(1-\\hat{p})}{n}}$",
        "$t\\frac{s}{\\sqrt{n}}\\sqrt{1-\\frac{n}{N}}$",    "$tN\\frac{s}{\\sqrt{n}}\\sqrt{1-\\frac{n}{N}}$")
ci <- c("$\\bar{x} \\pm t\\frac{s}{\\sqrt{n}}$", "$\\hat{p} \\pm z\\sqrt{\\frac{\\hat{p}(1-\\hat{p})}{n}}$", "$\\bar{x} \\pm t\\frac{s}{\\sqrt{n}}\\sqrt{1-\\frac{n}{N}}$",    "$N\\bar{x} \\pm tN\\frac{s}{\\sqrt{n}}\\sqrt{1-\\frac{n}{N}}$")
swr <- c("A","A","B","B")
d <- data.frame(Parameter = parameter, Estimate = estimator, 
  se = se, me = me, ci = ci, swr = swr)
d <- d[c(2,1,3,4),]
rownames(d) <- NULL
names(d)[2] <- "Point Estimate"
names(d)[3] <- "Standard Error"
names(d)[4] <- "Margin of Error"
names(d)[5] <- "Confidence Interval"
names(d)[6] <- "Design"
ktbl(d)
```

Design A: Sampling *with replacement*, an infinite number of observational units, or $N$ is much larger than $n$. 

Design B: Sampling *without replacement*. 

## Assumptions when Using these Equations

We are making certain *assumptions* about the *sampling distribution* of a point estimate when using the equations above. We assume that the statistic is *unbiased*, that we can compute (approximately) the correct *standard error*, and that the *shape of the sampling distribution is approximately normal*.

### The Mean of the Statistic

A statistic is **unbiased** if *its mean equals the parameter being estimated*, otherwise the statistic is said to be **biased**. 

Sources of bias:

1. **Sampling bias**: Failure to account for the fact that some units are more or less likely to be included in the sample. 

2. **Non-response bias**: Failure to observe some observational units that were intended to be observed.

3. **Response bias**: Errors in observation/measurement of the variable of interest.  

In surveys, *sampling bias* can be eliminated through the use of a *random sampling* design (aka "probability sampling" or "scientific sampling"). One example is what is known as a **simple random sampling** design where *every possible sample of $n$ units is equally likely*. Our formulas assume that the survey sampling design is simple random sampling. 

### The Standard Deviation of the Statistic

Recall that the *standard deviation of a statistic* is also called the **standard error**, which is used to compute the *margin of error*, the *confidence interval*, and *test statistics* (which we will discuss after the third examination).

The *design* can affect the standard error (e.g., sampling with versus without replacement).

### The Shape of the Statistic's Sampling Distribution

To have a sampling distribution that is (approximately) normal in shape, we need *at least one* of the following two conditions:

1. The population distribution is (approximately) normal in shape.

2. The sample size ($n$) is sufficiently large.

#### Shape of the Sampling Distribution of $\hat{p}$

1. The population distribution is *never* normal in shape. Why?

2. The sampling distribution is "sufficiently large" if both $n\hat{p} \ge 15$ and  $n(1-\hat{p}) \ge 15$. We can explore [why](http://digitalfirst.bfwpub.com/stats_applet/stats_applet_20_ciprop.html) this is necessary.

**Example**: A researcher obtains 100 seeds from a distributor. She plants those seeds and finds that 70 out of the 100 seeds germinate. She wants to estimate the probability that a seed obtained form this distributor will germinate using a confidence interval. Is the sample size large enough? What if she obtained 10 seeds and observed that 7 out of 10 germinated?

\vspace{2cm}

#### Shape of the Sampling Distribution of $\bar{x}$

1. The population distribution *might* be approximately normal in shape.

2. The sampling distribution will be approximately normal if $n$ is large enough.

**Example**: Here is a sample of $n$ = `r nrow(faithful)` observations of the eruption duration of Old Faithful. Would we conclude that the *population distribution* is approximately normal? Would we conclude that the *sampling distribution* is approximately normal? Why or why not?
```{r, fig.height = 3.25}
p <- ggplot(faithful, aes(x = round(eruptions,1))) 
p <- p + geom_dotplot(binwidth = 0.05)
p <- p + labs(x = "Eruption Duration (min)")
p <- p + theme_classic() + noyaxis
plot(p)
```
