---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("10-01-2022","12-01-2022","14-01-2022") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 1
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction and the "big picture"](lecture-01-12-2022.html).

`r i <- i + 1; d[i]` [Distributions, graphical depictions of distributions (i.e., histograms and dot plots), and summary measures](lecture-01-14-2022.html). A set of [homework problems](homework-01.html) on describing distributions in now available.

***

`r i <- i + 1; d[i]` **Holiday**.

`r i <- i + 1; d[i]` [Summary measures --- continued, cumulative distributions, and box plots](lecture-01-19-2022.html). A set of [homework problems](homework-02.html) on these topics is now available.

`r i <- i + 1; d[i]` [Normal/bell-shaped distributions](lecture-01-21-2022.html).
 
***

`r i <- i + 1; d[i]` [Outlier identification, margin of error](lecture-01-24-2022.html). A set of [homework problems](homework-03.html) on normal distributions and outliers is now available. The [study guide](e1-studyguide.html) for the first examination is also now available.

`r i <- i + 1; d[i]` [Margin of error --- continued, statistical significance, explanatory and response variables, types of studies](lecture-01-26-2022.html).

`r i <- i + 1; d[i]` **First examination**.
 
***

`r i <- i + 1; d[i]` [Introduction to probability and random variables](lecture-01-31-2022.html).

`r i <- i + 1; d[i]` [Computing probabilities from discrete and continuous probability distributions](lecture-02-02-2022.html). A set of [homework problems](homework-04.html) on probability distributions is now available.

`r i <- i + 1; d[i]` [Deriving sampling distributions](lecture-02-04-2022.html). A set of [homework problems](homework-05.html) on deriving sampling distributions is now available.
 
***

`r i <- i + 1; d[i]` [Sampling distributions of $\bar{x}$ and $\hat{p}$](lecture-02-07-2022.html).

`r i <- i + 1; d[i]` [Sampling distributions and the central limit theorem](lecture-02-09-2022.html). A set of [homework problems](homework-06.html) concerning the sampling distributions of $\bar{x}$ and $\hat{p}$ is now available. Also the [study guide](e2-studyguide.html) for the second examination is now available.

`r i <- i + 1; d[i]` [Sampling distribution of $s^2$ and examples of sampling distributions of $\bar{x}$ and $\hat{p}$](lecture-02-11-2022.html).
 
***

`r i <- i + 1; d[i]` **Second examination**.

`r i <- i + 1; d[i]` [Introduction to confidence intervals](lecture-02-16-2022.html). 

`r i <- i + 1; d[i]` [Confidence intervals for $\mu$ and $p$](lecture-02-18-2022.html). A set of [homework problems](homework-07.html) on confidence intervals for $\mu$ and $p$ is now available. 
 
***

`r i <- i + 1; d[i]` **Holiday**.

`r i <- i + 1; d[i]` [Sampling without replacement and estimating totals](lecture-02-23-2022.html). This lecture is recorded and that [recording is now available](https://vandalsuidaho-my.sharepoint.com/:v:/g/personal/trjohns_uidaho_edu/EROpH2Afal5Bo2686EUhavQBH8XgTLeBHAvmKgzUqu8Obw?e=LAGW0m). A set of [homework problems](homework-08.html) on these topics is available.

`r i <- i + 1; d[i]` **No class**.
 
***

`r i <- i + 1; d[i]` [Choosing a sample size](lecture-02-28-2022.html). A set of [homework problems](homework-09.html) on these topics is now available.

`r i <- i + 1; d[i]` [Assumptions](lecture-03-02-2022.html). The [study guide](e3-studyguide.html) for the third examination is now available.

`r i <- i + 1; d[i]` [Randomized response method](lecture-03-04-2022.html).
 
***

`r i <- i + 1; d[i]` **Third examination**.

`r i <- i + 1; d[i]` [Introduction to statistical tests](lecture-03-09-2022.html).

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Fourth examination**.

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Fifth examination**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Sixth examination**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

***

`r i <- i + 1; d[i]` **Final examination, 10:15 AM to 12:15 PM**.


