---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("23-08-2021","25-08-2021","27-08-2021") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 0
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction, samples versus populations, statistics versus parameters, description versus inference, and the "big picture"](lecture-08-23-2021.html).

`r i <- i + 1; d[i]` [Distributions and their description](lecture-08-25-2021.html). A [set of homework problems](homework-01.html) on describing distributions is now available.

`r i <- i + 1; d[i]` [Summary measures --- continued, cumulative distributions, percentiles, and box plots](lecture-08-27-2021.html). A set of [homework problems](homework-02.html) on cumulative distributions, percentiles, and box plots is now available.
 
***

`r i <- i + 1; d[i]` [Normal/bell-shaped distributions, the empirical rule, $z$-scores](lecture-08-30-2021.html). A set of [homework problems](homework-03.html) on normal distributions and outliers (to be discussed in the next lecture) is now available. 

`r i <- i + 1; d[i]` [Outlier identification, margin of error](lecture-09-01-2021.html). The [study guide](e1-studyguide.html) for the first examination is now available.

`r i <- i + 1; d[i]` [Margin of error, statistical significance, explanatory and response variables, types of studies](lecture-09-03-2021.html).
 
***

`r i <- i + 1; d[i]` Labor Day.

`r i <- i + 1; d[i]` **First examination**.

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Second examination**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Third examination**.

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Fourth examination**.

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Fifth examination**. 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` Fall recess.

`r i <- i + 1; d[i]` Fall recess.

`r i <- i + 1; d[i]` Fall recess.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Sixth examination**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 2; d[i]` **Final examination**, 8:00 to 10:00 AM.


