---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("23-08-2021","25-08-2021","27-08-2021") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 0
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction, samples versus populations, statistics versus parameters, description versus inference, and the "big picture"](lecture-08-23-2021.html).

`r i <- i + 1; d[i]` [Distributions and their description](lecture-08-25-2021.html). A [set of homework problems](homework-01.html) on describing distributions is now available.

`r i <- i + 1; d[i]` [Summary measures --- continued, cumulative distributions, percentiles, and box plots](lecture-08-27-2021.html). A set of [homework problems](homework-02.html) on cumulative distributions, percentiles, and box plots is now available.
 
***

`r i <- i + 1; d[i]` [Normal/bell-shaped distributions, the empirical rule, $z$-scores](lecture-08-30-2021.html). A set of [homework problems](homework-03.html) on normal distributions and outliers (to be discussed in the next lecture) is now available. 

`r i <- i + 1; d[i]` [Outlier identification, margin of error](lecture-09-01-2021.html). The [study guide](e1-studyguide.html) for the first examination is now available.

`r i <- i + 1; d[i]` [Margin of error, statistical significance, explanatory and response variables, types of studies](lecture-09-03-2021.html).
 
***

`r i <- i + 1; d[i]` Labor Day.

`r i <- i + 1; d[i]` **First examination**.

`r i <- i + 1; d[i]` [Introduction to probability and random variables](lecture-09-10-2021.html). A set of [homework problems](homework-04.html) on working with probability distributions is now available. You should be able to do the first section concerning means and standard deviations of random variables after this lecture. The lecture on Monday will cover what you need to know for the rest of this homework assignment. 
 
***

`r i <- i + 1; d[i]` [Computing probabilities from probability distributions](lecture-09-13-2021.html). 

`r i <- i + 1; d[i]` [Deriving sampling distributions](lecture-09-15-2021.html). A set of [homework problems](homework-05.html) on deriving distributions is now available.

`r i <- i + 1; d[i]` [Sampling distributions of $\bar{x}$ and $\hat{p}$](lecture-09-17-2021.html). A set of [homework problems](homework-06.html) on sampling distributions of $\bar{x}$ and $\hat{p}$ is now available.
 
***

`r i <- i + 1; d[i]` [Central limit theorem, estimation using $\bar{x}$ and $\hat{p}$](lecture-09-20-2021.html). The [study guide](e2-studyguide.html) for the second examination is now available.

`r i <- i + 1; d[i]` [The unbiasedness of $s^2$, review problems concerning the sampling distributions of $\bar{x}$ and $\hat{p}$](lecture-09-22-2021.html).

`r i <- i + 1; d[i]` **Second examination**.
 
***

`r i <- i + 1; d[i]` [Sampling distributions, error of estimation, and an introduction to confidence intervals](lecture-09-27-2021.html).

`r i <- i + 1; d[i]` [Confidence intervals for $\mu$ and $p$, confidence level](lecture-09-29-2021.html). A set of [homework problems](homework-07.html) on confidence intervals for $\mu$ and $p$ are now available.

`r i <- i + 1; d[i]` [Estimating $\mu$ and $\tau$ when sampling without replacement with known and finite $N$](lecture-10-01-2021.html). A set of [homework problems](homework-08.html) on these topics is now available. 
 
***

`r i <- i + 1; d[i]` [Choosing a sample size](lecture-10-04-2021.html). A set of [homework problems](homework-09.html) on choosing a sample size is now available.

`r i <- i + 1; d[i]` [Assumptions](lecture-10-06-2021.html). The [study guide](e3-studyguide.html) for the third examination is now available.

`r i <- i + 1; d[i]` [The randomized response method](lecture-10-08-2021.html). 
 
***

`r i <- i + 1; d[i]` **Third examination**.

`r i <- i + 1; d[i]` [Introduction to statistical tests](lecture-10-13-2021.html).

`r i <- i + 1; d[i]` [p-values, statistical tests concerning $p$](lecture-10-15-2021.html). A set of [homework problems](homework-10.html) on statistical tests concerning $p$ is now available.
 
***

`r i <- i + 1; d[i]` [Statistical tests concerning $\mu$](lecture-10-18-2021.html). A set of [homework problems](homework-11.html) on statistical tests concerning $\mu$ is now available.

`r i <- i + 1; d[i]` [Composite null hypothesis, the relationship between confidence intervals and tests, the sign test](lecture-10-20-2021.html). A set of [homework problems](homework-12.html) on these topics is now available.

`r i <- i + 1; d[i]` [Type I and Type II errors in statistical tests](lecture-10-22-2021.html). The [study guide](e4-studyguide.html) for the fourth examination is now available. 
 
***

`r i <- i + 1; d[i]` [Power](lecture-10-25-2021.html). 

`r i <- i + 1; d[i]` **Fourth examination**.

`r i <- i + 1; d[i]` [Independent versus dependent samples, inferences for $p_1 - p_2$](lecture-10-29-2021.html). A set of [homework problems](homework-13.html) on inferences for $p_1-p_2$ is now available.
 
***

`r i <- i + 1; d[i]` [Inferences for $\mu_1 - \mu_2$](lecture-11-01-2021.html). A set of [homework problems](homework-14.html) on inferences for $\mu_1 - \mu_2$ is now available.

`r i <- i + 1; d[i]` [Independent versus dependent samples --- revisited, the alternative standard error for inferences for $\mu_1-\mu_2$](lecture-11-03-2021.html). 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Fifth examination**. 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` Fall recess.

`r i <- i + 1; d[i]` Fall recess.

`r i <- i + 1; d[i]` Fall recess.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Sixth examination**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 2; d[i]` **Final examination**, 8:00 to 10:00 AM.


