---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("09-01-2023","11-01-2023","13-01-2023") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", 
  month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 1
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction and the "big picture"](lecture-01-11-2023.html).

`r i <- i + 1; d[i]` [Describing distributions](lecture-01-13-2023.html). The [first homework set](homework-01.html) on describing distributions is now available. 

***

`r i <- i + 1; d[i]` **Holiday**

`r i <- i + 1; d[i]` [Summary measures continued --- measures of variability, cumulative distributions, percentiles, box plots](lecture-01-18-2023.html). The [second homework set](homework-02.html) that concerns cumulative distributions, percentiles, and box plots is now available.

`r i <- i + 1; d[i]` [Normal-shaped distributions, *z*-scores](lecture-01-20-2023.html).
 
***

`r i <- i + 1; d[i]` [Identifying outliers, margin of error](lecture-01-23-2023.html). The [third homework set](homework-03.html) is now available. The [study guide](e1-studyguide.html) for the first examination is also now available.

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Exam 1**
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Exam 2**

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Holiday**

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Exam 3**
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Spring Recess**

`r i <- i + 1; d[i]` **Spring Recess**

`r i <- i + 1; d[i]` **Spring Recess**
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]`  

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Exam 4**

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Exam 5**

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Exam 6**
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]`

*** 

Tuesday, May 9. **Final Examination, 10:15 to 12:15**
