---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("09-01-2023","11-01-2023","13-01-2023") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", 
  month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 1
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction and the "big picture"](lecture-01-11-2023.html).

`r i <- i + 1; d[i]` [Describing distributions](lecture-01-13-2023.html). The [first homework set](homework-01.html) on describing distributions is now available. 

***

`r i <- i + 1; d[i]` **Holiday**

`r i <- i + 1; d[i]` [Summary measures continued --- measures of variability, cumulative distributions, percentiles, box plots](lecture-01-18-2023.html). The [second homework set](homework-02.html) that concerns cumulative distributions, percentiles, and box plots is now available.

`r i <- i + 1; d[i]` [Normal-shaped distributions, *z*-scores](lecture-01-20-2023.html).
 
***

`r i <- i + 1; d[i]` [Identifying outliers, margin of error](lecture-01-23-2023.html). The [third homework set](homework-03.html) is now available. The [study guide](e1-studyguide.html) for the first examination is also now available.

`r i <- i + 1; d[i]` [Margin of error --- continued, statistical significance, explanatory and response variables, types of studies](lecture-01-25-2023.html).

`r i <- i + 1; d[i]` **Exam 1**
 
***

`r i <- i + 1; d[i]` [Introduction to probability, random variables, and probability distributions](lecture-01-30-2023.html).

`r i <- i + 1; d[i]` [Computing probabilities from probability distributions](lecture-02-01-2023.html). A [homework assignment](homework-04.html) on working with probability distributions is now available.

`r i <- i + 1; d[i]` [Deriving sampling distributions](lecture-02-03-2023.html). A [homework assignment](homework-05.html) on deriving sampling distributions is now available.
 
***

`r i <- i + 1; d[i]` [Sampling distributions of $\bar{x}$ and $\hat{p}$](lecture-02-06-2023.html). A [set of homework problems](homework-06.html) concerning the sampling distributions of $\bar{x}$ and $\hat{p}$ is now available.

`r i <- i + 1; d[i]` [The central limit theorem, estimation](lecture-02-08-2023.html). The [study guide](e2-studyguide.html) for the second examination is now available.

`r i <- i + 1; d[i]` [Estimation and sampling distributions](lecture-02-10-2023.html). 
 
***

`r i <- i + 1; d[i]` **Exam 2**

`r i <- i + 1; d[i]` [Sampling distributions, estimation, and an introduction to confidence intervals](lecture-02-15-2023.html). 

`r i <- i + 1; d[i]` [Confidence intervals for $p$ and $\mu$](lecture-02-17-2023.html). A [set of homework problems](homework-07.html) on confidence intervals for $p$ and $\mu$ is now available.
 
***

`r i <- i + 1; d[i]` **Holiday**

`r i <- i + 1; d[i]` [Sampling with versus without replacement and estimating totals](lecture-02-22-2023.html). A [set of homework problems](homework-08.html) on this topic is now available.

`r i <- i + 1; d[i]` [Choosing a sample size](lecture-02-24-2023.html). A [set of homework problems](homework-09.html) on sample size selection is now available.
 
***

`r i <- i + 1; d[i]` [Assumptions](lecture-02-27-2023.html). The [study guide](e3-studyguide.html) for the third examination is now available.

`r i <- i + 1; d[i]` [The randomized response method](lecture-03-01-2023.html). 

`r i <- i + 1; d[i]` **Exam 3**
 
***

`r i <- i + 1; d[i]` [Introduction to statistical tests](lecture-03-06-2023.html). 

`r i <- i + 1; d[i]` [*p*-values, statistical tests concerning $p$](lecture-03-08-2023.html). A [set of homework problems](homework-10.html) on statistical tests concerning $p$ is now available.

`r i <- i + 1; d[i]` [Statistical tests concerning $\mu$](lecture-03-10-2023.html). A [set of homework problems](homework-11.html) on statistical tests concerning $\mu$ is now available.
 
***

`r i <- i + 1; d[i]` **Spring Recess**

`r i <- i + 1; d[i]` **Spring Recess**

`r i <- i + 1; d[i]` **Spring Recess**
 
***

`r i <- i + 1; d[i]` [Composite hypotheses, the relationship between tests and confidence intervals, and the sign test](lecture-03-20-2023.html). A [set of homework problems](homework-12.html) on these topics is now available.

`r i <- i + 1; d[i]` [Type I and Type II errors in statistical tests](lecture-03-22-2023.html). The [study guide](e4-studyguide.html) for the fourth examination is now available.

`r i <- i + 1; d[i]` [Power](lecture-03-24-2023.html). 
 
***

`r i <- i + 1; d[i]` **Exam 4**

`r i <- i + 1; d[i]` [Independent versus dependent samples, inferences for $p_1-p_2$](lecture-03-29-2023.html). A [set of homework problems](homework-13.html) concerning inferences for $p_1-p_2$ is now available.

`r i <- i + 1; d[i]` [Inferences for $\mu_1-\mu_2$](lecture-03-31-2023.html). A [set of homework problems](homework-14.html) concerning inferences for $\mu_1-\mu_2$ is now available.
 
***

`r i <- i + 1; d[i]` [Independent versus dependent samples, an alternative standard error for $\bar{x}_1-\bar{x}_2$](lecture-04-03-2023.html).

`r i <- i + 1; d[i]` [Causal inference](lecture-04-05-2023.html).
 
`r i <- i + 1; d[i]` [Survey sampling designs](lecture-04-07-2023.html). The [study guide](e5-studyguide.html) for the fifth examination is now available.
 
***

`r i <- i + 1; d[i]` [Misconceptions and limitations of statistical tests](lecture-04-10-2023.html).

`r i <- i + 1; d[i]` **Exam 5**

`r i <- i + 1; d[i]` [Goodness-of-fit tests](lecture-04-14-2023.html). 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` Class meeting in Albertson (ALB) 204.

`r i <- i + 1; d[i]` No class.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Exam 6**
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]`

*** 

Tuesday, May 9. **Final Examination, 10:15 to 12:15**
